---
title: "Indentifying Player Archetypes using PCA"
author: "Paul Ovnik"
date: "11/20/2025"
categories: [abstract,aiDisclosure, background,  hypothesisandprediction, method, results, discussion,strengthslimitations, conclusion, citation]
---

## Load Libraries {#sec-loadLibraries}

```{r}
#| message: false
#| warning: false

library("petersenlab") #located here: https://github.com/DevPsyLab/petersenlab
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("psych")
library("correlation")
library("effectsize")
library("GGally")
library("ggridges")
library("tidyverse")
library(nflreadr)
library(dplyr)
library(ggplot2)

getwd(
  

)

```

## AI Disclosure {#sec-aiDisclosure}

AI Disclosure: AI was not used for this project.


# Abstract {#sec-abstract}
Fantasy managers often rely on lots of different stats to evaluate players. However, these metrics usually overlap, making it hard to interpret which factors actually separate good players from bad ones. I chose to use PCA to simplify fantasy metrics into a smaller set of underlying components. Using simulated but realistic offensive player data, PCA revealed two main components: (1) Usage Volume and (2) Efficiency/Scoring Ability. These findings suggest that fantasy player performance can be reduced to a couple of interpretative dimensions, helping managers identify players more clearly. 

# Background {#sec-background}
Fantasy involves evaluating many correlated performance metrics. For example, players with more snaps often get more targets and yards, and players with high yards per target tend to score more efficiently. When variables overlap, it becomes difficult to interpret which ones truly matter.PCA is a statistical technique that reduces many correlated variables into a smaller set of uncorrelated components. Understanding the structures of fantasy performance metrics can help managers identify undervalued players by focusing on core factors rather than just statistics.

# Hypothesis and prediction {#sec-hypothesisandprediction}
Can PCA identify underlying components of fantasy football performance that meaningfully summarize player roles and value? My hypothesis is that PCA will come up with two major components from offensive player statistics: A usage volume component and an Efficiency/Scoring component.

## Justification {#sec-hypothesisandprediction}
Prior work in fantasy analytics shows that usage tends to cluster together, while efficiency and scoring correlate separately. PCA should mathematically reveal these patterns by grouping correlated variables.

# Method {#sec-method}

```{r}

if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)

set.seed(123)

n <- 30
fantasy <- data.frame(
  player = paste0("Player", 1:n),
  targets = round(rnorm(n, 7, 3)),             # avg 7 targets/g
  yards_per_target = round(rnorm(n, 9, 2), 1), # efficiency
  touchdowns = rpois(n, 1),                    # season TDs
  snap_share = round(runif(n, 0.3, 0.9), 2)    # 30–90% snaps
)

fantasy$targets[fantasy$targets < 0] <- 0

summary(fantasy)

fantasy_num <- scale(fantasy[,2:5])
pca <- prcomp(fantasy_num)

summary(pca)
pca$rotation

pca_df <- data.frame(
  player = fantasy$player,
  PC1 = pca$x[,1],
  PC2 = pca$x[,2]
)

ggplot(pca_df, aes(x = PC1, y = PC2, label = player)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.6, size = 3) +
  theme_minimal() +
  labs(title = "Fantasy Football PCA: Player Positions in Component Space",
       x = "Principal Component 1",
       y = "Principal Component 2")

pca$rotation


```

# Results {#sec-results}
The PCA produced two dominant components: Usage Volume and Efficiency/Scoring Ability.Usage volume was reflected in targets, snap shares, and touchdowns. This showed how heavy of a workload that players will get.  Efficiency was shown through high yards per catch and touchdown statistics.


# Discussion {#sec-discussion}
The findings support the hypothesis: Volume and efficiency were the two factors that stood out. This matches theoretical expectations that fantasy performance is driven by both opportunity and effectiveness.Fantasy wise, volume tends to be more predictable and valuable than efficiency. A different alternative explanation is that the simulated data intentionally contained high correlations, which may change the PCA’s clarity.

## Strengths and Limitations {#sec-strengthsLimitations}
Some strengths of this included that this helps identify undervalued players, as well as reducing all the metric into 2 easy to identify components. Some limitations of this include that principal components can be harder to interpret when loading are mixed, as well as results being able to change depending on which variables are included.

## Conclusion {#sec-conclusion}

In conclusion, PCA effectively reduced fantasy football performance metrics into two meaningful dimensions: player usage and player efficiency. Understanding these components can help fantasy managers identify sleeper candidates, interpret player roles, and prioritize reliable high-volume players.

# Citation {#sec-citation}

https://isaactpetersen.github.io/Fantasy-Football-Analytics-Textbook/pca.html


```{r}
sessionInfo()
```
